Placing the spotted T Tauri star LkCa 4 on an HR diagram

### 2016-08-15 document/ms.tex
Version of the paper GJH and MGS will review together in preparation to send to coauthors

In the previous section, we established that the high resolution optical and near-IR spectra of LkCa 4 may be explained by a two-temperature photosphere.  In this section, we use testable predictions from this fit to demonstrate that the fit reasonably matches spectral features and their rotational modulation.  We adopt the best-fit two temperature model to the IGRINS spectrum, with components of 2750 K covering 80\% of the visible stellar surface and 4200 K covering 20\% of the stellar surface.  The IGRINS spectrum occurred when LkCa 4 had an estimated brightness of $V_{\rm est}=12.84$ mag.  These parameters are selected  as roughly the median of the range of acceptable temperatures and filling factors from \S 3, blindly and without any adjustments to attempt to improve fits to the SED or broadband spectra.

Figures~\ref{fig:sed}-\ref{fig:lores} compares synthetic spectra from the two-component photosphere to the SED and flux-calibrated spectra of LkCa 4.  The only free parameters in this comparison are the luminosity and extinction, which are both scaled to match the spectrum as described below.  The optical-IR SED is obtained from estimating photometry from \citet{grankin08} during the 2MASS epoch.  The SED also includes mid-IR photometry from Spitzer/IRAC \citep{hartmann05}, without adjusting for epoch.  In the spectroscopic comparison, some minor discrepancies between the optical and near-IR spectra may be introduced because the spot coverage changed in the $\sim 5$ hrs between observations.  The synthetic spectrum is obtained from the Phoenix models, as in \S 3, and is extended beyond the longest wavelength (5 $\mu$m) for calculating the bolometric luminosity.

The synthetic models match the full spectrum reasonably well.  When scaled separately to the red-optical and near-IR spectra, the synthetic models match even better (Figure \ref{fig:lores_twopanel}).  However, single temperature photospheres (two shown in Figure \ref{fig:lores_twopanel}) fail to reproduce large spectral features.   The $3900$ K component fits the TiO bands reasonably well at $<7400$ \AA\ but fails to reproduce molecular bands at longer wavelengths.  The $3500$ K component hotter component suffers from the opposite problem, yielding molecular bands at short wavelengths that are too deep.  A two temperature photosphere with the parameters from the best-fit calculated in \S 3 fits both the molecular bands in the optical and near-IR spectra.

The features in the SED and in spectra should be rotationally modulated as the filling factors of the hot and cool components change.    In \S 4.1, we first use our fit to the IGRINS spectrum to convert the $V$-band brightness to the filling factor of the two components.  We subsequently calculate the color changes expected based on this spot coverage and compare these results to historical data (\S 4.2), and confirm our basic approach by demonstrating that TiO band depths vary with rotation (\S 4.3).

\begin{figure*}
 \centering
 %\includegraphics[trim=1.6cm 12.6cm 2.2cm 2.4cm, clip=true, width=0.60\textwidth]{figures/vband_spot_2750k}
\caption{The $V$-band magnitude in 2014--2015, converted into fill factor for the cool component.  The optical brightness depends mostly on the hot component.  If we fix a 75\% filling factor, as measured in the IGRINS spectrum, to $V=12.83$ at the time of the observation, then the $V$-band amplitude  corresponds to filling factors of 67--83\%.  The factor of $\sim 2$ change in visible surface area of the hot component, from 33\% to 17\%, is required to produce the $\delta V=0.6$. \todo{Update with all available photometry or the derived multi-term fit} }
\label{fig:vband_spot}
\end{figure*}

\begin{figure*}
 \centering
 \begin{tabular}{ll}
 %\includegraphics[trim=2.6cm 13.0cm 2.8cm 3.2cm, clip=true, width=0.45\textwidth]{figures/bv_pec13} 
    &
  %  \includegraphics[trim=2.6cm 13.0cm 2.8cm 2.2cm, clip=true, width=0.45\textwidth]{figures/vr_pec13}
    \end{tabular}
\caption{The observed optical colors of LkCa 4 from \citet{grankin08}, compared with predictions from converting $V$-band brightness to a cool spot filling factor and subsequently calculating colors from main sequence colors and bolometric corrections of \citet{kenyon95} (purple lines).}
\label{fig:colors}
\end{figure*}

@@

\begin{figure*}
 \centering
 %\includegraphics[trim=2.1cm 3.0cm 1cm 7.7cm, clip=true, width=0.70\textwidth]{figures/lores_panels_2750k}
 \caption{Top:  The low-resolution optical/near-IR spectrum of LkCa 4 obtained from Palomar/DBSP and APO/Triplespec on 30 December 2008 (black), compared to a synthetic spectrum of a two temperature photosphere (purple).  The inset shows that the 3000 K (red, 70\% fill factor) and 4200 K (blue, 30\% fill factor) components contribute equally to the near-IR spectrum, but the 4200 K component dominates the blue emission.  The synthetic spectrum is reddened by $A_V=0.4$ mag and scaled to the observed $J$-band spectrum. Bottom:  The low-resolution optical (left) and near-IR (right) spectrum of LkCa 4, compared with a 3900 K photosphere (blue), a 3500 K photosphere, and the two temperature photosphere that best fit the IGRINS spectrum.  The synthetic spectra are scaled separately to the optical spectrum at 0.75 $\mu$m and to the near-IR spectrum at 1.5 $\mu$m.  Warm photospheres accurately reproduce molecular bands at $0.7$ $\mu$m but fail to fit the spectral features at longer wavelengths.  Cooler photospheres predict molecular bands at $<0.7$ $\mu$m that are much deeper than observed.  The two temperature photosphere accurately fits spectral features in the optical and near-IR.}
 \label{fig:lores}
\end{figure*}

@@

The relative contributions from the hot and cool components change as LkCa 4 rotates, leading to rotational modulation of colors.  We model this rotational modulation using the main sequence colors compiled by \citet{pecaut13}, which should have less contribution from spots than their pre-main sequence colors.  The Cousins $R$ photometry from \citet{pecaut13} is converted to Johnson $R$ for comparison with the \citet{grankin08} photometry, following the Grankin et al.~prescription obtained from \citet{landolt82}.  Photometry from synthetic spectra produced by Phoenix and BT-Settl models cannot be used here because the predicted $B-V$ colors are much bluer than observed.  Offsets between model and observed colors are attributed to extinction and are discussed in \S 5.1.

@@

The correlation between $V$ and $V-R$ indicates that the star becomes redder when the cool spot is more dominant.  While this relationship is easily understood, the correlation predicted in our simple model is much weaker than the observed relationship.  Most likely, our two temperature model for the photosphere is overly simplistic, since the $V-R$ could could be reproduced with some contribution from intermediate temperature.

@@

\begin{figure}
 \centering
 %\includegraphics[trim=3.1cm 13cm 1.0cm 3.6cm, clip=true, width=0.49\textwidth]{figures/phot1_2750k}
\caption{The SED of LkCa 4 (red) compared with the two-temperature spectrum obtained from the best fit to the IGRINS spectra (black).}
\label{fig:sed}
\end{figure}

@@

The depths of TiO and other molecular bands are common diagnostics of spectral type (and therefore temperature) in optical spectra \citep[e.g.][]{kirkpatrick91}.  In a 2-temperature photosphere cool enough for molecules to form, the depth of observed TiO bands will depend on the fractional coverage of the components.  In this subsection, we demonstrate that the TiO band depths vary with spot coverage using 12 epochs of CFHT/ESPaDOnS spectra obtained over 14 days ($\sim 4$ rotation periods) by \citet{donati14}.

@@

@@

Multi-temperature photospheres may resolve some of the long-standing problems related to discrepancies between observed and model colors of young stars.  Low mass young stars are redder than expected in the near-IR \citep[e.g.][]{tottle15}.  As a consequence, extinctions are high when measured by comparing near-IR colors to colors expected for optical spectral types.  For LkCa 4, the excess $E(J-K)$ color leads to an extinction of $1.04$ mag for a 4200 K photosphere and \citep{pecaut13} colors.  This extinction is consistent with the color and SED analysis of \citet{furlan06}, which adopted a K7 spectral type, and is significantly higher than extinctions of $A_V=0.69$ measured from optical photometry \citep{kenyon95} and $0.35$ mag measured by comparing optical spectra to young spectral templates \citep{herczeg14}.

An accurate extinction estimate from the full SED is challenging because of a strong dependence on the areal distribution of temperatures in the photosphere, which in our work is simplified to a two temperature fit.  However, isolating specific colors constrains the possible range of extinctions.  
The $B-V$ color is stable and not affected by cooler components, and may therefore be compared with the expected colors for the hotter (4100 K) temperature.  
The average $B-V=1.41$ mag color from \citet{grankin08} has a color excess $E(B-V)=0.09$ mag relative to the empirical color expected for a main sequence star and $E(B-V)=0.20$ mag for colors of pre-main sequence stars, based on the temperature-color tables of \citep{pecaut13}.  The main sequence colors are probably most 
appropriate in this analysis because they are more free of spots.  
The $B-V$ color 
excess leads to an extinction $A_V=0.28\pm0.15$ mag, or $0.62\pm0.15$ mag relative to 
pre-main sequence colors, with the listed uncertainty calculated for similar estimates with a hot component of 4000 and 4200 K.   The $V-R_J$ color excess leads to $A_V=0.48$ mag, which should be less affected by gravity (based on $V-I_C$ colors in Pecaut et al.) but may be overestimated because the affects of the cool component are underestimated in our two-temperature model (as seen in Fig.~X).
In the near-IR, the estimated $V=12.61$ mag at the time of the 2MASS observations leads to a template (main sequence) color 
$J-K_s=0.87$ mag, with a range of $\sim 0.05$ mag for acceptable solutions with other 
temperatures and covering fractions. The observed $J-K=0.93$ mag leads to $A_V=0.35$ mag, 
with a range of $0.29$ mag to account for the range of other acceptable photospheric solutions.  
The template near-IR color is more robust than optical colors to differences 
in the main sequence/pre-main sequence temperature scales because the pre-main 
sequence $J-K$ color is redder at 4200 K and bluer at 2800 K.  Based on this 
analysis, an $A_V=0.3$ mag is adopted throughout this paper.

The near-IR extinctions of LkCa 4 may be overestimated if emission is generated from a composite spectrum but comparison colors come from the hotter component.  This mismatch likely explains why near-IR extinctions are often higher than optical extinctions for young stars \citep[see analysis in][]{herczeg14}.  Our results tentatively support the lower extinctions obtained at optical wavelengths, although the template colors used to calculate extinction are uncertain and will need revision.  In the \citet{pecaut13} conversions, late K spectral types lead to lower temperatures for pre-main sequence than main-sequence stars, and at the same temperature the B-V colors are redder.  The extinction to LkCa 4 and similar objects depends on whether these differences are caused by spots or gravity.  
Uncertainty in colors may be solved with self-consistent comparisons to young stars with similar spectral types, if spot coverage is also similar.

@@

The apparent effective temperature in any single epoch would depend on the relative fractions of the hot and cool spot.  Decreasing the cool spot size to 60\% would increase the effective temperature to 3670 K, while increasing the spot size to 85\% would decrease the effective temperature to 3370 K.  This wide range of possible observed effective temperatures at any single epoch, when accounting for the visible contributions of both hot and cool components, underscores the difficulty in placing young stars on HR diagrams.  Since the Hayashi tracks are nearly vertical at this temperature and age, the range in effective temperatures that would be measured in any single epoch corresponds to a mass range of $\sim 0.27--0.4$ $M_\odot$.

@@

Our confusion in colors, spectral type/effective temperatures, and extinctions for TTSs is dominated by ignoring spots and is not a fundamental problem with either photospheric models or with spectral type-effective temperature scales.  Any single spectral type from either optical or near-IR spectra will provide a misleading estimate of the effective temperature.

@@

>>>>

\begin{figure}
 \centering
 \includegraphics[trim=3.1cm 13cm 1.0cm 3.6cm, clip=true, width=0.49\textwidth]{figures/phot_2750}
\caption{The SED of LkCa 4 (red) compared with the two-temperature spectrum obtained from the best fit to the IGRINS spectra (black).}
\label{fig:sed}
\end{figure}

\begin{figure*}
 \centering
 \includegraphics[trim=2.1cm 3.0cm 1cm 7.7cm, clip=true, width=0.70\textwidth]{figures/lores_panels_2750k}
 \caption{Top:  The low-resolution optical/near-IR spectrum of LkCa 4 obtained from Palomar/DBSP and APO/Triplespec on 30 December 2008 (black), compared to a synthetic spectrum of a two temperature photosphere (purple).  The inset shows that the 2750 K (red, 70\% fill factor) and 4100 K (blue, 30\% fill factor) components contribute equally to the near-IR spectrum, but the 4100 K component dominates the blue emission.  The synthetic spectrum is reddened by $A_V=0.4$ mag and scaled to the observed $J$-band spectrum. Bottom:  The low-resolution optical (left) and near-IR (right) spectrum of LkCa 4, compared with a 3900 K photosphere (blue), a 3500 K photosphere, and the two temperature photosphere that best fit the IGRINS spectrum.  The synthetic spectra are scaled separately to the optical spectrum at 0.75 $\mu$m and to the near-IR spectrum at 1.5 $\mu$m.  Warm photospheres accurately reproduce molecular bands at $0.7$ $\mu$m but fail to fit the spectral features at longer wavelengths.  Cooler photospheres predict molecular bands at $<0.7$ $\mu$m that are much deeper than observed.  The two temperature photosphere accurately fits spectral features in the optical and near-IR.}
 \label{fig:lores}
\end{figure*}

In the previous section, we established that the high resolution optical and near-IR spectra of LkCa 4 may be explained by a two-temperature photosphere.  In this section, we use testable predictions from this fit to demonstrate that the fit reasonably matches spectral features and their rotational modulation.  We adopt the best-fit two temperature model to the IGRINS spectrum, with components of 2750 K covering 80\% of the visible stellar surface and 4100 K covering 20\% of the stellar surface.  The IGRINS spectrum occurred when LkCa 4 had an estimated brightness of $V_{\rm est}=12.84$ mag.  These parameters are selected without any adjustments to attempt to improve fits to the SED or broadband spectra.

Figures~\ref{fig:sed}-\ref{fig:lores} compares synthetic spectra from the two-component photosphere to the SED and flux-calibrated spectra of LkCa 4.  The only free parameters in this comparison are the luminosity and extinction, which are both scaled to match the spectrum (see \S 5.1-5.2).  The optical-IR SED is obtained from estimating photometry from \citet{grankin08} during the 2MASS epoch, with $V\sim12.61$ mag.  The SED also includes mid-IR photometry from Spitzer/IRAC \citep{hartmann05}, without adjusting for epoch.  In the spectroscopic comparison, some minor discrepancies between the optical and near-IR spectra may be introduced because the spot coverage changed in the $\sim 5$ hrs between observations ($\Delta V=0.08$ mag).  The synthetic spectrum is obtained from the Phoenix models, as in \S 3, and is extended beyond the longest wavelength (5 $\mu$m) for calculating the bolometric luminosity.

The synthetic models match the full spectrum reasonably well.  When scaled separately to the red-optical and near-IR spectra, the synthetic models match even better (bottom panels of Figure \ref{fig:lores}).  However, single temperature photospheres fail to reproduce large spectral features.   The $3900$ K component fits the TiO bands reasonably well at $<7400$ \AA\ but fails to reproduce molecular bands at longer wavelengths.  The $3500$ K component hotter component suffers from the opposite problem, yielding molecular bands at short wavelengths that are too deep.  A two temperature photosphere with the parameters from the best-fit calculated in \S 3 reproduces the TiO bands, the hump in the $H$-band, and the jump in flux at the long-wavelength end of the $J$-band.

The spectral features should be rotationally modulated as the filling factors of the hot and cool components change.    In \S 4.1, we first use our fit to the IGRINS spectrum to convert the $V$-band brightness to the filling factor of the two components.  We subsequently calculate the color changes expected based on this spot coverage and compare these results to historical data (\S 4.2), and confirm our basic approach by demonstrating that TiO band depths vary with rotation (\S 4.3).

\begin{figure*}
 \centering
 \includegraphics[trim=1.6cm 12.6cm 2.2cm 2.4cm, clip=true, width=0.60\textwidth]{figures/vband_spot_2750k}
\caption{The $V$-band magnitude in 2014--2015, converted into fill factor for the cool component.  The optical brightness depends mostly on the hot component.  If we fix a 75\% filling factor, as measured in the IGRINS spectrum, to $V=12.83$ at the time of the observation, then the $V$-band amplitude  corresponds to filling factors of 67--83\%.  The factor of $\sim 2$ change in visible surface area of the hot component, from 33\% to 17\%, is required to produce the $\delta V=0.6$. \todo{Update with all available photometry or the derived multi-term fit} }
\label{fig:vband_spot}
\end{figure*}

\begin{figure*}
 \centering
 \begin{tabular}{ll}
 \includegraphics[trim=2.6cm 13.0cm 2.8cm 3.2cm, clip=true, width=0.45\textwidth]{figures/bv_pec13} 
    &
   \includegraphics[trim=2.6cm 13.0cm 2.8cm 2.2cm, clip=true, width=0.45\textwidth]{figures/vr_pec13}
    \end{tabular}
\caption{The observed optical colors of LkCa 4 from \citet{grankin08}, compared with predictions from converting $V$-band brightness to a cool spot filling factor and subsequently calculating colors from main sequence colors and bolometric corrections of \citet{kenyon95} (purple lines).}
\label{fig:colors}
\end{figure*}

@@

@@

The relative contributions from the hot and cool components change as LkCa 4 rotates, leading to rotational modulation of colors.  We model this rotational modulation using the main sequence colors \citep[compiled by][]{pecaut13}, which should have less contribution from spots than  pre-main sequence colors.  The Cousins $R$ photometry from \citet{pecaut13} is converted to Johnson $R$ for comparison with the \citet{grankin08} photometry, following the color transformation prescribed by \citet{landolt82} and applied by Grankin et al.  Photometry from synthetic spectra produced by Phoenix and BT-Settl models cannot be used here because the predicted $B-V$ colors are much bluer than observed.  Offsets between model and observed colors are attributed to extinction and are discussed in \S 5.1.

@@

The correlation between $V$ and $V-R$ indicates that the star becomes redder when the cool spot has a higher filling factor.  Our simple model predicts that the correlation should be much weaker than is measured.  Most likely, our two temperature model for the photosphere is overly simplistic, since the $V-R$ could could be reproduced with some contribution from intermediate temperature.

@@

@@

The depths of TiO and other molecular bands are common diagnostics of spectral type and therefore temperature in optical spectra \citep[e.g.][]{kirkpatrick91}.  In a 2-temperature photosphere cool enough for molecules to form, the depth of observed TiO bands will depend on the fractional coverage of the components.  In this subsection, we demonstrate that the TiO band depths vary with spot coverage using 12 epochs of CFHT/ESPaDOnS spectra obtained over 14 days ($\sim 4$ rotation periods) by \citet{donati14}.

@@

Some overview?

@@

Multi-temperature photospheres may resolve some of the long-standing problems related to discrepancies between observed and model colors of young stars.  Low mass young stars are redder than expected in the near-IR, when compared with expectations from optical spectral types \citep[e.g.][]{tottle15}.  For LkCa 4, the excess $E(J-K)$ color leads to an extinction of $1.04$ mag for a 4200 K photosphere and \citep{pecaut13} colors.  This extinction is consistent with the color and SED analysis of \citet{furlan06}, which adopted a K7 spectral type, and is significantly higher than extinctions of $A_V=0.69$ measured from optical photometry \citep{kenyon95} and $0.35$ mag measured by comparing optical spectra to young spectral templates \citep{herczeg14}.

An accurate extinction estimate from the full SED is challenging because of a strong dependence on the areal distribution of temperatures in the photosphere, which in our work is simplified to a two temperature fit.  However, isolating specific colors constrains the possible range of extinctions.  
The $B-V$ color is stable and not affected by cooler components, and may therefore be compared with the expected colors for the hotter (4100 K) temperature.  
The average $B-V=1.41$ mag color from \citet{grankin08} has a color excess $E(B-V)=0.09$ mag relative to the empirical color expected for a main sequence star and $E(B-V)=0.20$ mag for colors of pre-main sequence stars, based on the temperature-color tables of \citep{pecaut13}.  The main sequence colors are probably most 
appropriate in this analysis because they are more free of spots.  
The $B-V$ color 
excess leads to an extinction $A_V=0.28\pm0.15$ mag, or $0.62\pm0.15$ mag relative to 
pre-main sequence colors.   The listed uncertainties are calculated from extinction estimates relative to colors of 4000 and 4200 K photospheres.
The $V-R_J$ color excess leads to $A_V=0.48$ mag, which should be less affected by gravity (based on $V-I_C$ colors in Pecaut et al.) but is likely to be overestimated because the cool component contributes to the $R_J$-band emission, as seen in Fig.~\ref{fig:colors}.
In the near-IR, the estimated $V=12.61$ mag at the time of the 2MASS observations leads to a template (main sequence) color 
$J-K_s=0.87$ mag, with a range of $\sim 0.05$ mag for acceptable solutions with other 
temperatures and covering fractions. The observed $J-K=0.93$ mag leads to $A_V=0.35$ mag, 
with a range of $0.29$ mag to account for the range of other acceptable photospheric solutions.  
The template near-IR color is more robust than optical colors to differences 
in the main sequence/pre-main sequence temperature scales because the pre-main 
sequence $J-K$ color is redder at 4100 K and bluer at 2750 K {\bf GJH: (just redo for Pec13 young colors)}.  Based on this 
analysis, an $A_V=0.3$ mag is adopted throughout this paper.

The near-IR extinctions of LkCa 4, and perhaps other TTSs, will be overestimated if emission is generated from a multi-temperature spectrum but template colors are based on only the hotter component.  This mismatch likely explains why near-IR extinctions are often higher than optical extinctions for young stars \citep[see analysis in][]{herczeg14}.  Our results tentatively support the lower extinctions obtained at optical wavelengths, although the template colors used to calculate extinction are uncertain and will need revision.  Differences between main sequence and pre-main sequence temperature scales and colors in the \citet{pecaut13} tables are introduced by a combination of spots, in addition to any color differences caused by gravity differences.
Uncertainty in colors may be solved with self-consistent comparisons to young stars with similar spectral types, if spot coverage is also similar.

@@

The apparent effective temperature in any single epoch depends on the relative fractions of the hot and cool spot.  Decreasing the cool spot size to 60\% increases the effective temperature to 3670 K, while increasing the spot size to 85\% decreases to the effective temperature to 3370 K.  This wide range of possible observed effective temperatures at any single epoch, when accounting for the visible contributions of both hot and cool components, underscores the difficulty in placing young stars on HR diagrams.  Since the Hayashi tracks are nearly vertical at this temperature and age, this range in effective temperatures corresponds to a mass range of $\sim 0.27-0.4$ $M_\odot$  {\bf (GJH: all numbers need revision for 2750 K)}.

@@

Discrepancies in colors, spectral type/effective temperatures, and extinctions for LkCa 4, and perhaps more broadly for TTSs, are dominated by ignoring spots and are not caused by fundamental problems in either photospheric models or in spectral type-effective temperature scales.

@@

0 New

Reply
